# Data

## Technical description
The primary data source for this project is the INEI, Peru's official statistics agency. INEI collects a wide array of data, including population statistics and sample surveys. These datasets are typically available in formats like Excel and are updated periodically, with some reports published quarterly or annually. For instance, the Encuesta Nacional de Hogares (ENAHO) provides detailed information on household living conditions, income, expenses, and employment. 

Data collection: 
We will need to process the data to create a master table from the url 
https://proyectos.inei.gob.pe/microdatos/. 

Data collection for our especific anlaysis is conducted at the household level using a sample survey approach. The survey is carried out quarterly due to the complexity and budget requirements that a full census of the entire population would entail. This sample-based approach ensures statistically representative data across Peru’s official geographic divisions.

Period: 
Second quarter of 2025 (most updated information)

Note: 
The column labels and the page are in Spanish. We will translate the columns. 

The ENAHO survey gathers the following information: 

- Housing and household characteristics
- Individual and demographic characteristics
- Education
- Health
- Food and beverage expenditures
- Social programs (by household members)
- Governance, democracy, and transparency 

The data is collected at the household level, which can be aggregated by block, zone, district, province, department, and country, reflecting the official geographic divisions of Peru.

SES levels:
In addition to ENAHO, a supplementary dataset classifying Lima metropolitan districts by socioeconomic level (A, B, C, D, E) has been incorporated. This dataset shows the composition of each district across these five SES levels and has been used to cross-reference districts with other variables in our analysis.

https://apeim.com.pe/wp-content/uploads/2024/01/APEIM-Informe-de-Niveles-Socioeconomicos-2023-2024-Version-WEB.pdf


## Missing value analysis

```{r}
data <- read.csv('master_left.csv')
```

For treatment of missing values, please refer to the Readme. AS
In this part, we worked with the data without any treatment.
(To make my point, I'm choosing some columns; however, we have more than ~2189 columns
to analyze, which we did in Python. However, I will show you how we should proceed here.

First, we analyze the number of missing values per column. For this, there is this plot.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

vars <- c(
  "shock_business_failure",
  "shock_job_loss",
  "shock_none",
  "shock_natural_disaster",
  "shock_crime_theft",
  "shock_household_head_abandonment",
  "shock_illness_accident",
  "bribery_last_12m",
  "last_approved_grade",
  "last_approved_year",
  "last_approved_institution_type",
  "water_access_daily",
  "water_is_potable"
)

df_selected <- data |> select(any_of(vars))

miss_long <- df_selected %>%
  mutate(row_id = row_number()) %>%
  mutate(across(-row_id, as.character)) %>%   # <- makes all columns same type
  pivot_longer(
    cols = -row_id,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(missing = if_else(is.na(value), "yes", "no"))

ggplot(miss_long, aes(x = variable, y = factor(row_id), fill = missing)) +
  geom_raster(color = "grey85", linewidth = 0.15) +
  scale_fill_manual(values = c("no" = "#2c105c", "yes" = "#fde725")) +
  labs(title = "Missingness map (selected variables)",
       x = "variable", y = "row", fill = "missing") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

We analyzed the type of variables and the number of unique values, and we noticed an interesting pattern. The word 'pase' in Spanish is the same as 'pass' in English. We reviewed the manual instructions and found that 'pass' indicates a non-response in the census questionnaire. Then we replace every 'pass' with NA, and we have the following graph.

```{r}
df_selected <- df_selected %>%
  mutate(across(
    everything(),
    ~ if_else(. %in% c("pase", "Pase"), NA, .)
  ))

miss_long <- df_selected %>%
  mutate(row_id = row_number()) %>%
  mutate(across(-row_id, as.character)) %>%   # <- makes all columns same type
  pivot_longer(
    cols = -row_id,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(missing = if_else(is.na(value), "yes", "no"))


ggplot(miss_long, aes(x = variable, y = factor(row_id), fill = missing)) +
  geom_raster(color = "grey85", linewidth = 0.15) +
  scale_fill_manual(values = c("no" = "#2c105c", "yes" = "#fde725")) +
  labs(title = "Missingness map (selected variables)",
       x = "variable", y = "row", fill = "missing") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

As we can observe, now the columns that were supposed to be 100% without nulls have nulls. Additionally, some of them are in the same columns as the cases of 'shock_business_failure', 'shock_crime_theft', and ' shock_household_head_abandonment'.

After this analysis, we replicated for all the columns in our base (more info in the read me file )
